### 2022-03-07 bools/bools.tex
more

\looseness=-1
A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
extension of superposition for extensional Boolean-free higher-order logic. The extension removes the need to translate the
first two above mentioned features of higher-order logic. Kotelnikov et al.\
\cite{kotelnikov-15-fool,kotelnikov-16-fool} extended the language of
first-order logic to support the third feature of higher-order logic that
requires translation. They described two approaches: One based on calculus-level
treatment of Booleans and the other, which requires no changes to the calculus,
based on preprocessing.

@@

\looseness=-1
Rules that extend the two orthogonal approaches form the basis of our support
for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
implemented rules that are inspired by the ones used in the
higher-order provers Leo-III \cite{sb-21-leo3} and Satallax
\cite{cb-12-satallax}, such as elimination of Leibniz equality, primitive
instantiation and treatment of choice operator \cite{pa-01-classical-ty-thy}. We
have also designed new rules that use higher-order unification to resolve
Boolean formulas that are hoisted to literal level, delay clausification of non-atomic
literals, reason about formulas under $\lambda$-binders, and many others. Even
though the rules we use are inspired by the ones of refutationally complete
higher-order provers, we do not guarantee completeness of our extension of
$\lambda$-superposition.

@@

\looseness=-1
A signature is a quadruple $(\Sigmaty, \Vty, \Sigma, \VV)$ where $\Sigmaty$ is a
set of type constructors, $\Vty$ is a set of type variables and $\Sigma$ and
$\VV$ are sets of constants and term variables, respectively. We require nullary
type constructor $o$, as well as binary constructor $\rightarrow$
to be in $\Sigmaty$. A type $\tau, \upsilon$ is either a type variable $\alpha \in
\Vty$ or of the form $\kappa(\tau_1, \ldots \tau_n)$ where $\kappa$ is an
$n$-ary type constructor. We write $\kappa$ for $\kappa()$, $\tau \rightarrow
\upsilon$ for $\rightarrow(\tau, \upsilon)$, and we drop parentheses to shorten 
$\tau_1 \rightarrow (\cdots \rightarrow (\tau_{n-1} \rightarrow \tau_n) \cdots)$ into $\tau_1 \rightarrow \cdots \rightarrow
\tau_n$. Each symbol in $\Sigma$ is
assigned a type declaration of the form $\forallty{\tuplen{\alpha}} \tau$ where all variables
occurring in $\tau$ are among $\tuplen{\alpha}$.

@@

\newcommand{\eqneq}{\mathrel{\dot{\eq}}}
These two rules hoist Boolean subterms $f$ to the literal level; therefore,
some results of \infname{Cases} and \infname{CasesSimp} will have literals of the form $f \eq \itrue$ (or
$f \noteq \itrue$) where $f$ is not an atom. This introduces the need for the rule
called immediate clausification (\infname{IC}):
%
$$ \namedsimp{EC}{C}{D_1 \; \cdots \; D_m} $$
%
We say that a clause is \emph{standard} if all of its literals are of the form $s \eqneq t$,
where $s$ and $t$ are not Booleans or of the form $f \eqneq \itrue$, where the head of $f$
is not a logical symbol and $\eqneq$ denotes $\eq$ or $\not\eq$. The rule \infname{IC}
is applicable if clause $C = L_1 \llor
\cdots \llor L_n$ is not standard.
The resulting clauses $\tuple{D}{m}$ represent
the result of clausification of the formula $\iforall\,\tuple{x}{}. \; L_1 \ior
\cdots \ior L_n$ where $\tuple{x}{}$ are all free variables of $C$.
Using Boolean extensionality, Zipperposition's clausification
algorithm treats Boolean equality as equivalence (i.e., it replaces
$\mathop{\eq}\typeargs{o}$ with $\iimplies$).

\looseness=-1
An advantage of leaving nested Booleans unmodified is that the prover will be able
to prove some problems containing them without using the prolific rules described
above. For example, given two clauses $\cst{f} \, (\cst{p} \, x
\iimplies \cst{p} \, y) \eq \cst{a}$ and $\cst{f} \, (\cst{p} \,
\cst{a} \iimplies \cst{p} \, \cst{b}) \noteq \cst{a}$, the empty clause can
easily be derived without the above rules. A disadvantage of this approach
is that the proving process will periodically be interrupted by expensive calls
to the clausifier.
\pagebreak[2]

@@

\subsection{Higher-Order Considerations}
\label{subsect:bool:core}
\looseness=-1
To achieve refutational completeness of higher-order resolution and similar
calculi it is necessary to instantiate variables with result type $o$,
\emph{predicate variables}, with arbitrary formulas
\cite{as-18-phd,pa-01-classical-ty-thy}. Fortunately, we can approximate the
formulas using a complete set of logical symbols (e.g., $\inot$, $\iforall$, and
$\iand$). Since such an approximation is not only necessary for completeness of
some calculi, but very useful in practice, we implemented the \emph{primitive
instantiation} (PI) rule:
%
$$ \namedinference{PI} {C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq}
\, t} {\substcl{\{ P
\mapsto f \}}{C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq} \, t}  } $$
%
where $P$ is a free variable of
the type $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow o$. 
Choosing a different $f$ that instantiates $P$, we can balance between
explosiveness of approximating a complete set of logical symbols and
incompleteness of pragmatic approaches. We borrow the notion of imitation from
higher-order unification jargon (Sect.~\ref{sec:unif:the-unification-procedure}) %\cite{OUR UNIFICATION PAPER}:
, and we say
that the term $\lam{\tuple{x}{m}}{\cst{f} \, (Y_1 \, \tuple{x}{m}) \cdots (Y_n
\, \tuple{x}{m}) }$ is an \emph{imitation} of constant $\cst{f} : \tau_1
\rightarrow \cdots \rightarrow \tau_n \rightarrow \tau$ for some variable $z$ of type $\nu_1
\rightarrow \cdots \rightarrow \nu_m \rightarrow \tau$. Variables $\tuplen{Y}$
are fresh free variables, where each $Y_i$ has the type $\nu_1 \rightarrow
\cdots \rightarrow \nu_m \rightarrow \tau_i$; variable $x_i$ is of type $\nu_i$.
\pagebreak[2]

\looseness=-1
Rule \infname{PI} was already implemented by Simon Cruanes in Zipperposition,
before we started our modifications. The rule has different modes that generate
sets of possible terms $f$ for $p: \tau_1 \rightarrow \cdots \rightarrow \tau_n
\rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$ where
$\star$ is an element of a set of logical constants $P = \{ \iand, \ior,
\mathop{\ieq}\typeargs{\alpha}, \inot, \iforall, \iexists \}$. Mode \emph{Full}
contains imitations (for $p$) of all elements of $P$. Mode $\emph{Pragmatic}$
contains imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
such that $i\ineq j$ and  $\tau_i = \tau_j$, it contains $\lam{\tuplen{x}}{x_i
\ieq x_j}$; if there exist indices $i,j$ such that $i \ineq j$, and $\tau_i =
\tau_j = o$, then it contains $\lam{\tuplen{x}}{x_i \iand x_j}$ and
$\lam{\tuplen{x}}{x_i \ior x_j}$; if for some $i$, $\tau_i = o$, then it
contains $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ contains
imitations of $\itrue$, $\ifalse$ and $\star$ (except for $\mathit{Imit_{\iforall\iexists}}$ which contains imitations
of both $\iforall$ and $\iexists$).

@@

\noindent\looseness=-1
Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are extensional
versions of superposition, equality resolution and equality factoring
(Sect.~\ref{sec:pre:rules}). By \infname{Abs} we denote the union of these three
rules. In each of the rules, $\sigma$ is a most general unifier of the types of
$s$ and $s'$, and $\textsf{dp}(s\sigma ,s' \sigma) = \{ (s_1, s'_1), \ldots,
(s_n, s'_n)\}$. All side conditions for extensional rules are the same as for
the standard rules, except that condition that $s$ and $s'$ are unifiable is
replaced by the condition that at least one $s_i$ is of functional type and that
$n>0$. This rule is easily extended to support Boolean extensionality by
requiring that at least one $s_i$ is of functional or type $o$, and adding the
condition ``$\textsf{dp}(f, g) = \{(f,g)\}$ if $f$ and $g$ are different
formulas'' to the definition of $\textsf{dp}$.

@@

Before we began our modifications, Zipperposition had a powerful rule that
recognizes clauses that contain variations of Leibniz equality and instantiates
them with native equality. This rule was designed by Simon Cruanes, and to the
best of our knowledge, it has not been documented so far. With his permission
we describe this rule as follows:
\pagebreak[2]
%
$$\namedinference{ElimPredVar}{P \, \overline{s}_n^1 \eq \itrue \llor \cdots \llor
    P \, \overline{s}_n^i \eq \itrue \llor P \, \overline{t}_n^1 \noteq \itrue \llor
    \cdots \llor P \, \overline{t}_n^j \noteq \itrue \llor C} 
    {\sigmacl{P \, \overline{s}_n^1
    \eq \itrue \llor \cdots \llor P \, \overline{s}_n^i \eq \itrue  \llor C}} $$
%
where $P$ is a free variable, $P$ does not occur in any $s_k^l$ or $t_k^l$, or
in $C$; $\sigma$ is defined as $\{ P \mapsto \lam{\tuplen{x}}{\pmb{\bigvee}_{k=1}^{j}
(\pmb{\bigwedge}_{l=1}^{n} x_l \eq t^k_l ) } \}$.

@@

The previous rule removes all variables occurring in disequations in one
attempt. We implemented two rules that behave more lazily, inspired by the ones present in Leo-III and
Satallax:
%
\begin{align*}
&\namedinference{ElimLeibniz$+$}
{P \, \tuplen{s} \eq \itrue \llor P \, \tuplen{t} \noteq \itrue \llor C}
{\sigmacl{s_i \eq t_i \llor C}}
&&
\namedinference{ElimLeibniz$-$}
{P \, \tuplen{s} \noteq \itrue \llor P \, \tuplen{t} \eq \itrue \llor C}
{\substcl{\sigma'}{s_i \eq t_i \llor C}}&
\end{align*}
%
\looseness=-1
where $P$ is a free variable, $P$ does not occur in $t_i$, $\sigma =
\{ P \mapsto \lam{\tuplen{x}}{x_i \eq t_i} \}$ and
$\sigma' = \{ P \mapsto \lam{\tuplen{x}}{\neg (x_i \eq t_i)} \}$. This
rule differs from \infname{ElimPredVar} in three ways. First, it acts on
occurrences of variables in both positive and negative literals. Second,
due to its simplicity, it usually does not require \infname{IC} as the following step.
Third, it imposes much weaker conditions on $P$. However, removing all negative
variables in one step might improve performance.  Coming back to example of the clause $C = P \, \cst{a}_1 \, \cst{a}_2 \eq \itrue \llor P \,
\cst{b}_1 \, \cst{b}_2 \not\eq \itrue \llor P \, \cst{c}_1 \, \cst{c}_2 \not\eq
\itrue$, we can apply \infname{ElimLeibniz$+$} using the substitution $\sigma = \{ P \mapsto \lam{xy}{x \eq \cst{b_1}} \}$
to obtain the clause $C' = \cst{a}_1 \eq \cst{b}_1 \llor \cst{a}_1 \not\eq \cst{c}_1$.

% \subsection{Additional Rules}
% \label{subsect:bool:core}

% Zipperposition's unification algorithm 
% %\cite{OUR UNIFICATION PAPER}
% uses flattened representation of terms with logical operators $\land$ and $\lor$
% for heads to unify terms that are not unifiable modulo $\alpha\beta\eta$-equivalence, but
% are unifiable modulo associativity and commutativity of $\land$ and $\lor$. Let
% $\diamond$ denote either $\land$ or $\lor$. When the unification algorithm is given
% two terms $\diamond \, \tuplen{s}$ and $\diamond \, \tuplen{t}$, where neither
% of $\tuplen{s}$ nor $\tuplen{t}$ contain duplicates, it performs the
% following steps: First, it removes all terms that appear in both
% $\tuplen{s}$ and $\tuplen{t}$ from the two argument tuples.
% %Let $\tuple{s}{m}$ and $\tuple{t}{m}$ denote the results of previous step.
% Next, the remaining terms are sorted first by their head term and then their weight. Finally,
% an attempt is made to unify sorted lists pairwise.
% %
% As an example, consider the problem of unifying the pair $\big({\land} \, (\cst{p} \,
% \cst{a}) \; (\cst{q} \, (\cst{f} \, \cst{a})), \; {\land} \; (\cst{q} \, (\cst{f} \,
% \cst{a})) \; (r \, (\cst{f} \, (\cst{f} \, \cst{a})))\bigr)$ where $r$ is a free
% variable. If the arguments of $\land$ are simply sorted as described above, we would
% try to unify $\cst{p} \, \cst{a}$ with $\cst{q} \, (\cst{f} \, \cst{a})$, and
% fail to find a unifier. However, by removing term  $\cst{q} \, (\cst{f} \,
% \cst{a})$ from the argument lists, we will be left with the problem
% $(\cst{p} \, \cst{a},  r \, (\cst{f} \, (\cst{f} \, \cst{a})))$ which has a unifier.

>>>>

\looseness=-1
A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
extension of superposition for extensional Boolean-free higher-order logic. The extension removes the need to translate
$\lambda$-binders and functional extensionality. Kotelnikov et al.\
\cite{kotelnikov-15-fool,kotelnikov-16-fool} extended the language of
first-order logic to support the third feature of higher-order logic that
requires translation. They described two approaches: One based on calculus-level
treatment of Booleans and the other, which requires no changes to the calculus,
based on preprocessing.

@@

\looseness=-1
Rules that extend the two orthogonal approaches form the basis of our support
for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
implemented rules that are inspired by the ones used in the
higher-order provers Leo-III \cite{sb-21-leo3} and Satallax
\cite{cb-12-satallax}, such as elimination of Leibniz equality, primitive
instantiation and treatment of choice operator \cite{pa-01-classical-ty-thy}. We
have also designed new rules that use higher-order unification to resolve
Boolean formulas that are hoisted to literal level, delay clausification of nonatomic
literals, reason about formulas under $\lambda$-binders, and many others. Even
though the rules we use are inspired by the ones of refutationally complete
higher-order provers, we do not guarantee completeness of our extension of
$\lambda$-superposition.

@@

\looseness=-1
A {\em signature} is a quadruple $(\Sigmaty, \Vty, \Sigma, \VV)$ where $\Sigmaty$ is a
set of type constructors, $\Vty$ is a set of type variables and $\Sigma$ and
$\VV$ are sets of constants and term variables, respectively. We require nullary
type constructor $o$, as well as binary constructor $\rightarrow$
to be in $\Sigmaty$. A type $\tau, \upsilon$ is either a type variable $\alpha \in
\Vty$ or of the form $\kappa(\tau_1, \ldots \tau_n)$ where $\kappa$ is an
$n$-ary type constructor. We write $\kappa$ for $\kappa()$, $\tau \rightarrow
\upsilon$ for $\rightarrow(\tau, \upsilon)$, and we drop parentheses to shorten 
$\tau_1 \rightarrow (\cdots \rightarrow (\tau_{n-1} \rightarrow \tau_n) \cdots)$ into $\tau_1 \rightarrow \cdots \rightarrow
\tau_n$. Each symbol in $\Sigma$ is
assigned a type declaration of the form $\forallty{\tuplen{\alpha}} \tau$ where all variables
occurring in $\tau$ are among $\tuplen{\alpha}$.

@@

\newcommand{\eqneq}{\mathrel{\dot{\eq}}}
These two rules hoist Boolean subterms $f$ to the literal level; therefore,
some results of \infname{Cases} and \infname{CasesSimp} will have literals of the form $f \eq \itrue$ (or
$f \noteq \itrue$) where $f$ is not an atom. This introduces the need for the rule
called immediate clausification (\infname{IC}):
%
$$ \namedsimp{EC}{C}{D_1 \; \cdots \; D_m} $$
%
We say that a clause is \emph{standard} if all of its literals are of the form $s \eqneq t$,
where $s$ and $t$ are not Booleans or of the form $f \eqneq \itrue$, where the head of $f$
is not a logical symbol and $\eqneq$ denotes $\eq$ or $\not\eq$. The rule \infname{IC}
is applicable if clause $C = L_1 \llor
\cdots \llor L_n$ is not standard.
The resulting clauses $\tuple{D}{m}$ represent
the result of clausification of the formula $\iforall\,\tuple{x}{}. \; L_1 \ior
\cdots \ior L_n$ where $\tuple{x}{}$ are all free variables of $C$.
Using Boolean extensionality, Zipperposition's clausification
algorithm treats Boolean equality as equivalence (i.e., it replaces
$\mathop{\eq}\typeargs{o}$ with $\iequiv$).

\looseness=-1
An advantage of leaving nested Booleans unmodified is that the prover will be able
to prove some problems containing them without using the prolific rules described
above. For example, given two clauses $\cst{f} \, (\cst{p} \, X
\iimplies \cst{p} \, Y) \eq \cst{a}$ and $\cst{f} \, (\cst{p} \,
\cst{a} \iimplies \cst{p} \, \cst{b}) \noteq \cst{a}$, the empty clause can
easily be derived without the above rules. A disadvantage of this approach
is that the proving process will periodically be interrupted by expensive calls
to the clausifier.
\pagebreak[2]

@@

\subsection{Higher-Order Considerations}
\label{subsect:bool:core}
\looseness=-1
To achieve refutational completeness of higher-order resolution and similar
calculi it is necessary to instantiate variables with result type $o$,
\emph{predicate variables}, with arbitrary formulas
\cite{as-18-phd,pa-01-classical-ty-thy}. Fortunately, we can approximate the
formulas using a complete set of logical symbols (e.g., $\inot$, $\iforall$, and
$\iand$). Since such an approximation is not only necessary for completeness of
some calculi, but very useful in practice, we implemented the \emph{primitive
instantiation} (PI) rule:
%
$$ \namedinference{PI} {C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq}
\, t} {\substcl{\{ P
\mapsto f \}}{C \llor \lam{\tuple{x}{}}{P \, \tuplen{s}} \, \dot{\eq} \, t}  } $$
%
where $P$ is a free variable of
the type $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow o$. 
Choosing a different $f$ that instantiates $P$, we can balance between
explosiveness of approximating a complete set of logical symbols and
incompleteness of pragmatic approaches. We borrow the notion of imitation from
higher-order unification jargon (Sect.~\ref{sec:unif:the-unification-procedure}), and we say
that the term $\lam{\tuple{x}{m}}{\cst{f} \, (Y_1 \, \tuple{x}{m}) \cdots (Y_n
\, \tuple{x}{m}) }$ is an \emph{imitation} of constant $\cst{f} : \tau_1
\rightarrow \cdots \rightarrow \tau_n \rightarrow \tau$ for some variable $Z$ of type $\nu_1
\rightarrow \cdots \rightarrow \nu_m \rightarrow \tau$. Variables $\tuplen{Y}$
are fresh free variables, where each $Y_i$ has the type $\nu_1 \rightarrow
\cdots \rightarrow \nu_m \rightarrow \tau_i$; variable $x_i$ is of type $\nu_i$.
\pagebreak[2]

\looseness=-1
Rule \infname{PI} was already implemented by Simon Cruanes in Zipperposition,
before we started our modifications. The rule has different modes that generate
sets of possible terms $f$ for $P: \tau_1 \rightarrow \cdots \rightarrow \tau_n
\rightarrow o$: \emph{Full}, \emph{Pragmatic}, and $\mathit{Imit}_{\star}$ where
$\star$ is an element of a set of logical constants $P = \{ \iand, \ior,
\mathop{\ieq}\typeargs{\alpha}, \inot, \iforall, \iexists \}$. Mode \emph{Full}
contains imitations (for $P$) of all elements of $P$. Mode $\emph{Pragmatic}$
contains imitations of $\inot$, $\itrue$ and $\ifalse$; if there exist indices $i,j$
such that $i\ineq j$ and  $\tau_i = \tau_j$, it contains $\lam{\tuplen{x}}{x_i
\ieq x_j}$; if there exist indices $i,j$ such that $i \ineq j$, and $\tau_i =
\tau_j = o$, then it contains $\lam{\tuplen{x}}{x_i \iand x_j}$ and
$\lam{\tuplen{x}}{x_i \ior x_j}$; if for some $i$, $\tau_i = o$, then it
contains $\lam{\tuplen{x}}{x_i}$. Mode $\mathit{Imit}_\star$ contains
imitations of $\itrue$, $\ifalse$ and $\star$ (except for $\mathit{Imit_{\iforall\iexists}}$ which contains imitations
of both $\iforall$ and $\iexists$).

@@

\noindent\looseness=-1
Rules \infname{AbsSup}, \infname{AbsER}, and \infname{AbsEF} are extensional
versions of superposition, equality resolution and equality factoring
(Sect.~\ref{sec:pre:rules}). By \infname{Abs} we denote the union of these three
rules. In each of the rules, $\sigma$ is a most general unifier of the types of
$s$ and $s'$, and $\textsf{dp}(\sigmaterm{s} ,\sigmaterm{s'}) = \{ (s_1, s'_1), \ldots,
(s_n, s'_n)\}$. All side conditions for extensional rules are the same as for
the standard rules, except that condition that $s$ and $s'$ are unifiable is
replaced by the condition that at least one $s_i$ is of functional type and that
$n>0$. This rule is easily extended to support Boolean extensionality by
requiring that at least one $s_i$ is of functional or type $o$, and adding the
condition ``$\textsf{dp}(f, g) = \{(f,g)\}$ if $f$ and $g$ are different
formulas'' to the definition of $\textsf{dp}$.

@@

Before we began our modifications, Zipperposition had a powerful rule that
recognizes clauses that contain variations of Leibniz equality and instantiates
them with native equality. This rule was designed by Simon Cruanes, and to the
best of our knowledge, it has not been documented so far. With his permission
we describe this rule as follows:
\pagebreak[2]
%
$$\namedinference{ElimPredVar}{P \, \overline{s}_n^1 \eq \itrue \llor \cdots \llor
    P \, \overline{s}_n^i \eq \itrue \llor P \, \overline{t}_n^1 \noteq \itrue \llor
    \cdots \llor P \, \overline{t}_n^j \noteq \itrue \llor C} 
    {\sigmacl{P \, \overline{s}_n^1
    \eq \itrue \llor \cdots \llor P \, \overline{s}_n^i \eq \itrue  \llor C}} $$
% \pagebreak[2]
where $P$ is a free variable, $P$ does not occur in any $s_k^l$ or $t_k^l$, or
in $C$; $\sigma$ is defined as $\{ P \mapsto \lam{\tuplen{x}}{\pmb{\bigvee}_{k=1}^{j}
(\pmb{\bigwedge}_{l=1}^{n} x_l \eq t^k_l ) } \}$.

@@

The previous rule removes all variables occurring in disequations in one
attempt. We implemented two rules that behave more lazily, inspired by the ones present in Leo-III and
Satallax:
%
\begin{align*}
&\namedinference{ElimLeibniz$+$}
{P \, \tuplen{s} \eq \itrue \llor P \, \tuplen{t} \noteq \itrue \llor C}
{\sigmacl{s_i \eq t_i \llor C}}
&&
\namedinference{ElimLeibniz$-$}
{P \, \tuplen{s} \noteq \itrue \llor P \, \tuplen{t} \eq \itrue \llor C}
{\substcl{\sigma'}{s_i \eq t_i \llor C}}&
\end{align*}
%
\looseness=-1
where $P$ is a free variable, $P$ does not occur in $t_i$, $\sigma =
\{ P \mapsto \lam{\tuplen{x}}{x_i \eq t_i} \}$ and
$\sigma' = \{ P \mapsto \lam{\tuplen{x}}{\neg (x_i \eq t_i)} \}$. This
rule differs from \infname{ElimPredVar} in three ways. First, it acts on
occurrences of variables in both positive and negative literals. Second,
due to its simplicity, it usually does not require \infname{IC} as the following step.
Third, it imposes much weaker conditions on $P$. However, removing all negative
variables in one step might improve performance.  Coming back to example of the clause $C = P \, \cst{a}_1 \, \cst{a}_2 \eq \itrue \llor P \,
\cst{b}_1 \, \cst{b}_2 \not\eq \itrue \llor P \, \cst{c}_1 \, \cst{c}_2 \not\eq
\itrue$, we can apply \infname{ElimLeibniz$+$} using the substitution $\sigma = \{ P \mapsto \lam{xy}{x \eq \cst{b}_1} \}$
to obtain the clause $C' = \cst{a}_1 \eq \cst{b}_1 \llor \cst{a}_1 \not\eq \cst{c}_1$.

\subsection{Additional Rules}
\label{subsect:bool:core}

Zipperposition's unification algorithm 
%\cite{OUR UNIFICATION PAPER}
uses flattened representation of terms with logical operators $\iand$ and $\ior$
for heads to unify terms that are not unifiable modulo $\alpha\beta\eta$-equivalence, but
are unifiable modulo associativity and commutativity of $\iand$ and $\ior$. Let
$\Diamond$ denote either $\iand$ or $\ior$. When the unification algorithm is given
two terms $\Diamond \, \tuplen{s}$ and $\Diamond \, \tuplen{t}$, where neither
of $\tuplen{s}$ nor $\tuplen{t}$ contain duplicates, it performs the
following steps: First, it removes all terms that appear in both
$\tuplen{s}$ and $\tuplen{t}$ from the two argument tuples.
%Let $\tuple{s}{m}$ and $\tuple{t}{m}$ denote the results of previous step.
Next, the remaining terms are sorted first by their head term and then their weight. Finally,
an attempt is made to unify sorted lists pairwise.
%
As an example, consider the problem of unifying the pair $\big({\iand} \,
(\cst{p} \, \cst{a}) \; (\cst{q} \, (\cst{f} \, \cst{a})), \; {\iand} \;
(\cst{q} \, (\cst{f} \, \cst{a})) \; (R \, (\cst{f} \, (\cst{f} \,
\cst{a})))\bigr)$ where $R$ is a free variable. If the arguments of $\iand$ are
simply sorted as described above, we would try to unify $\cst{p} \, \cst{a}$
with $\cst{q} \, (\cst{f} \, \cst{a})$, and fail to find a unifier. However, by
removing term  $\cst{q} \, (\cst{f} \, \cst{a})$ from the argument lists, we
will be left with the problem $(\cst{p} \, \cst{a},  R \, (\cst{f} \, (\cst{f}
\, \cst{a})))$ which has a unifier.

The winner of THF division of
CASC-27 \cite{gs-19-casc27}, Satallax \cite{cb-12-satallax}, has one
crucial advantage over Zipperposition: it is based on higher-order
tableaux, and as such it does not require formulas to be converted to clauses. The advantage of tableaux is that once it instantiates a variable with a
term, this instantiation naturally propagates through the whole formula. In
Zipperposition, which is based on higher-order superposition, the original formula
is clausified and instantiating a variable in a clause $C$ does not automatically instantiate
it in all clauses that are results of clausification of the same formula as $C$.
To mitigate this issue, we have created extensions of equality resolution and equality factoring
that take Boolean extensionality into account:
%
\begin{align*}
  & \namedinference{BoolER}
  { s \eq s' \llor C}
  {\sigmacl{C}}
  &&
  \namedinference{BoolEF$+-$}
  {P\,\tuplen{s} \eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}} \\
  & \namedinference{BoolEF$-+$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \eq \itrue \llor C}
  {\sigmacl{s' \eq \itrue \llor C}} &&
  \namedinference{BoolEF$-$$-$}
  {P\,\tuplen{s} \not\eq \itrue \llor s' \not\eq \itrue \llor C}
  {\sigmacl{s' \not\eq \itrue \llor C}}
\end{align*}
%
All side conditions except for the ones concerning the unifiability of terms are
as in the original equality resolution and equality factoring rules. In rule
\infname{BoolER}, $\sigma$ is a unifier of $ s $ and $ \neg s' $. In the $+-$
and $-+$ versions of $\infname{BoolEF}$, $\sigma$ unifies $ x \, \tuplen{s}
$ and $\neg s'$, and in the remaining version it unifies $ x \, \tuplen{s} $ and
$s'$. Intuitively, these rules bring Boolean (dis)equations in the appropriate
form for application of the corresponding base rules. It suffices to consider 
literals of the form $s \eq s'$ for \infname{BoolER} as Zipperposition rewrites
$s \iequiv t \eq \itrue$ and $\inot(s \iequiv t) \not\eq \itrue$ to $s \eq t$ (and does analogous
rewriting into $ s \not\eq t $).

Another approach to mitigate harmful effects of immediate clausification
is to delay it as long as possible. Following the approach by 
Ganzinger and Stuber \cite{gs-05-boolsup}, we represent every input formula $f$ as a unit clause $f \eq \itrue$ and use the following
delayed clausification (\infname{DC}) rules:

\vspace{2\jot}
\begin{tabular}{ccc}
  $\namedsimp{DC$_{\iand}$}
  { (f \iand g) \eq \itrue \llor C}
  { f \eq \itrue \llor C \quad g \eq \itrue \llor C }$ &
  $\namedsimp{DC$_{\ior}$}
  { (f \ior g) \eq \itrue \llor C}
  { f \eq \itrue \llor g \eq \itrue  \llor C }$ &
  $\namedsimp{DC$_{\iimplies}$}
  { (f \iimplies g) \eq \itrue \llor C}
  { f \not\eq \itrue \llor g \eq \itrue  \llor C } $ \\[2\jot]
  $\namedsimp{DC$_{\inot}$}
  { (\inot f) \eq \itrue \llor C}
  { f \not\eq \itrue \llor C}$ &
  $\namedsimp{DC$_{\iforall}$}
  { (\iforall x. \, f) \eq \itrue \llor C}
  { \{x \mapsto Y\}f \llor C }$ &
  $\namedsimp{DC$_{\iexists}$}
  { (\iexists x.\, f) \eq \itrue \llor C}
  { \{x \mapsto \cst{sk}\typeargs{\tuple{\alpha}{}} \, \tuplen{Y} \}f \llor C }$ 
  \\[2\jot]
  \multicolumn{3}{c}{$\namedsimp{DC$_\eq$}
  { f \eq g \llor C}
  { f \not\eq \itrue \lor g\eq\itrue \llor C \quad f \eq \itrue \lor g\not\eq\itrue \llor C}$}
\end{tabular}
\vspace{2\jot}

\looseness=-1
The rules described above are as given by Ganzinger and Stuber (adapted to
our setting), with the omission of rules for negative literals ($ f \not\eq
\itrue$), which are easy to derive and which can be found in their work \cite{gs-05-boolsup}.
In \infname{DC}$_\eq$ we require both $f$ and $g$ to be formulas and at least one of
them not to be $\itrue$. In \infname{DC}$_{\iforall}$, $Y$ is a fresh variable, and in
\infname{DC}$_{\iexists}$, $\cst{sk}$ is a fresh symbol and $\tuple{\alpha}{}$ and $\tuplen{Y}$ are all the type and term variables occurring
freely in $\iexists x.\, f$.

\looseness=-1
Naive application of the \infname{DC} rules can result in exponential blowup in
problem size. To avoid this, we rename formulas that have repeated
occurrences. We keep the count of all nonatomic formulas occurring as
either side of a literal. Before applying the \infname{DC} rule on a
clause $f \mathrel{\dot{\eq}} \itrue \llor C $, we check whether the number of $f$'s
occurrences exceeds the threshold $k$. If it does, based on the
polarity of the literal $f \mathrel{\dot{\eq}} \itrue$, we add the clause $\cst{p} \, \tuplen{Y}
\not\eq \itrue \llor f \eq \itrue$ (if the literal is positive) or $\cst{p} \, \tuplen{Y}
\eq \itrue \llor f \not\eq \itrue$ (if the literal is negative), where $\tuplen{Y}$ are
all free variables of $f$ and $\cst{p}$ is a fresh symbol. Then, we replace the
clause $f \mathrel{\dot\eq} \itrue \llor C$ by $ \cst{p} \, \tuplen{Y} \mathrel{\dot\eq} \itrue \llor
C$.

Before the number of occurrences of $f$ is checked, we first check (using a fast, incomplete matching algorithm) if there is a
formula $g$, for which definition was already introduced, such that $\sigmaterm{g} =
f$, for some substitution $\sigma$. This check can have three outcomes. First, if
the definition $\textsf{q} \, \overline{X}_n$ is already introduced for $g$ with the
polarity matching that of $f \, \dot{\eq} \, \itrue$, then $f$ is
replaced by $\sigmaterm{\cst{q} \, \tuplen{X}}$. Second, if the definition was introduced,
but with different polarity, we create the clause defining $g$ with the missing
polarity, and replace $f$ with $\sigmaterm{\cst{q} \, \tuplen{X}}$. Last, if the there is no renamed
formula $g$ generalizing $f$, then we perform the previously described check.

\looseness=-1
In addition to reusing names for formula definitions, we reuse the Skolem
symbols introduced by the \infname{DC}$_{\iexists}$ rule. When
\infname{DC}$_{\iexists}$ is applied to $f = \iexists x.\, f'$ we check if there is
a Skolem $\cst{sk}\typeargs{\tuple{\alpha}{m}} \, \tuplen{Y}$ introduced for a
formula $g = \iexists x.\, g'$, such that $\sigmaterm{g} = f$. If so, the symbol
$\cst{sk}$ is reused and $\iexists x.\, f'$ is replaced by $\{x \mapsto
(\cst{sk}\typeargs{\tuple{\alpha}{m}} \, \tuplen{Y})\sigma \}f'$. Renaming and
name reusing techniques are inspired by the VCNF algorithm described by Reger et
al.\cite{rsv-16-vcnf}.

% \namedinference{LC$_\eq$}
% { f \eq g \llor C}
% { f \not\eq \ifalse \llor g \eq \ifalse \llor C \quad f \eq \ifalse \llor g \not\eq \ifalse \llor C } &&
% \namedinference{LC$_\Rightarrow$}

% Rule \infname{BoolSimp} proved to be very useful in our initial experimentation
% as it normalizes many equivalent terms to the same Boolean term. Stronger normalization can be achieved if nested formulas
% are transformed into \emph{negation normal form} (\emph{NNF}). Formula is in negation normal form if it is built from atoms and their
% negations using only $\wedge$ and $\vee$ \cite{nw-01-small-cnf}. By
% adding
% \kern\abovedisplayskip
